---
title: 'Reproducible Research: Peer Assignment 2'
author: "Scott Hansen"
date: "Wednesday, March 04, 2016"
output: html_document
---
# Data Preparation
## Initial Settings

```{r setoptions, echo=FALSE}
  knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)

  setwd("E:\\Data Scientist\\Data Science Track\\Reproducible Reseearch\\Project2\\Reproducible-Research-Project-2")
```

```{r LoadPackages, results="hide"}

x<-c("R.utils", "ggplot2", "lubridate", "plyr", "gridExtra")

lapply(x, library, character.only=TRUE)

```

## Load Data
If the data is not already downloaded in the working directory, we will attempt to retrieve and unzip the file from the internet

```{rGetData, echo=TRUE}
if (!"repdata-data-StormData.csv.bz2" %in% dir(".")) {
    print("File Not Found... Attempting to download")
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "repdata-data-StormData.csv.bz2")
    bunzip2("repdata-data-StormData.csv.bz2", overwrite=T, remove=F)
}
```

### Load Variable
Load and investigate the data set.  Cache is turned on to improve performance.

```{r LoadData}
  if(!"StormData" %in% ls())
  {
    StormData<-read.csv("repdata-data-StormDatarepdata-data-StormData.csv", stringsAsFactors = FALSE)
  }
  dim(StormData)
  head(StormData, n=1)
  str(StormData)
```

## Investigate Data set
Histogram of storm events

```{r histStormEvents}
  StormData$Year<- as.numeric(year(mdy_hms(StormData$BGN_DATE)))
  hist(StormData$Year, breaks=61)

```
## Chosing the best subset of the data
It is unlikely that there were significantly more storms starting around 1990.  More likely record keeping improved; so the analysis will focus only on weather events after 1990
```{r reshapeData}
  StormData<-StormData[StormData$Year >= 1990, ]
  dim(StormData)
```

## Build Summary Statistics
### Fatalities
```{r Fatalities}
  Fatalities<-aggregate(FATALITIES~EVTYPE, data=StormData, sum)
  Fatalities<-head(arrange(Fatalities, Fatalities$FATALITIES, decreasing=TRUE), n=10)
```

### Injuries
```{r Injuries}
  Injuries<-aggregate(INJURIES~EVTYPE, data=StormData, sum)
  Injuries<-head(arrange(Injuries, Injuries$INJURIES, decreasing=TRUE), n=10)
  
```
### Economic Impact
The economic impact cannot be directly read from the data fram.  Based on the metadata about the data set, there is an exponential multiplier that must be applied.

* Summarize Property and Crop damage
* Eliminate Bad Exponential exponential multipliers

```{r EconomicImpact}

  PropDamage<- aggregate(PROPDMG~EVTYPE+PROPDMGEXP, data=StormData, sum)
  PropDamage<- PropDamage[PropDamage$PROPDMGEXP %in% c("K", "M", "B"),]

  CropDamage<- aggregate(CROPDMG~EVTYPE+CROPDMGEXP, data=StormData, sum)
  CropDamage<- CropDamage[CropDamage$CROPDMGEXP %in% c("K", "M", "B"),]

  for (i in 1:nrow(CropDamage))
    {

    if (CropDamage$CROPDMGEXP[i]=="K") {CropDamage$CROPDMG[i]= CropDamage$CROPDMG[i]*1000}
    if (CropDamage$CROPDMGEXP[i]=="M") {CropDamage$CROPDMG[i]= CropDamage$CROPDMG[i]*1000000}
    if (CropDamage$CROPDMGEXP[i]=="B") {CropDamage$CROPDMG[i]= CropDamage$CROPDMG[i]*1000000000}
    }
    
  for (i in 1:nrow(PropDamage))
    {

    if (PropDamage$PROPDMGEXP[i]=="K") {PropDamage$PROPDMG[i]= PropDamage$PROPDMG[i]*1000}
    if (PropDamage$PROPDMGEXP[i]=="M") {PropDamage$PROPDMG[i]= PropDamage$PROPDMG[i]*1000000}
    if (PropDamage$PROPDMGEXP[i]=="B") {PropDamage$PROPDMG[i]= PropDamage$PROPDMG[i]*1000000000}
    }

```
### Develop a Top 10 list for crop damage, property damage and total damage
* Create a summary based on event type
* Order the summarized data set and select the top 10 storm event types
* Print results

```{r EconSummary}
  CropDamage<-aggregate(CROPDMG~EVTYPE, data=CropDamage, sum)
  CropDamage<-head(arrange(CropDamage, CropDamage$CROPDMG, decreasing=TRUE), n=10)

  PropDamage<-aggregate(PROPDMG~EVTYPE, data=PropDamage, sum)
  PropDamage<-head(arrange(PropDamage, PropDamage$PROPDMG, decreasing=TRUE), n=10)

```
# Presentation of Results
This analysis will break the analysis of weatherevents into 2 part:

* Human cost
* Economic costs

## Presenting the human cost of weather
The top 10 weather events for fatalities are:
```{r presentFatalities}
  Fatalities
```
The top 10 weather events for Injuries are:
```{r presentInjuries}
  Injuries
```
This can be best presented in a bar graph

```{r graphHumanCost}
fatalitiesPlot <- ggplot(Fatalities, aes(reorder(EVTYPE, FATALITIES), FATALITIES)) +
  geom_bar(stat="identity") + 
    scale_y_continuous("# of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Weather Type") + 
    ggtitle("Total Fatalities by Weather Type\n Events in the U.S.\n from 1990 - 2011")
injuriesPlot <- ggplot(Injuries, aes(reorder(EVTYPE, INJURIES), INJURIES)) +
  geom_bar(stat="identity") + 
    scale_y_continuous("# of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Weather Type") + 
    ggtitle("Total Injuries by Weather Type\n Events in the U.S.\n from 1990 - 2011")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
```
Based on the above histograms, we find that excessive heat and tornado cause most fatalities; tornato causes most injuries in the United States from 1990 to 2011.


## Presenting the economic cost of weather
The top 10 weather events for Property damage are:
```{r presentProperty}
  PropDamage
```
The top 10 weather events for Crop damage are:
```{r presentCrops}
  CropDamage
```

This can be best presented in a bar graph

```{r graphEconomicCosts}
propertyPlot <- ggplot(PropDamage, aes(reorder(EVTYPE, PROPDMG), PROPDMG)) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+ 
  xlab("Weather Type") + ggtitle("Total Property Damage by\n Weather Type in\n the U.S. from 1990 - 2011")


cropPlot<- ggplot(CropDamage, aes(reorder(EVTYPE, CROPDMG), CROPDMG)) +
  geom_bar(stat="identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") + 
    xlab("Weather Type") + ggtitle("Total Crop Damage by \nWeather Type in\n the U.S. from 1990 - 2011")

grid.arrange(propertyPlot, cropPlot, ncol = 2)
```

Based on the above histograms, we find that flood and hurricane/typhoon cause most property damage; drought and flood causes most crop damage in the United States from 1990 to 2011.

# Conclusion
From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.