---
title: 'Reproducible Research: Peer Assignment 2'
author: "Scott Hansen"
date: "Wednesday, March 02, 2016"
output: html_document
---

## Initial Settings

```{r setoptions, echo=FALSE}
  knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)

  setwd("E:\\Data Scientist\\Data Science Track\\Reproducible Reseearch\\Project2\\Reproducible-Research-Project-2")
```

```{r LoadPackages, results="hide"}

x<-c("R.utils", "ggplot2", "lubridate", "plyr")

lapply(x, library, character.only=TRUE)

```

## Load Data
If the data is not already downloaded in the working directory, we will attempt to retrieve and unzip the file from the internet

```{rGetData, echo=TRUE}
if (!"repdata-data-StormData.csv.bz2" %in% dir(".")) {
    print("File Not Found... Attempting to download")
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "repdata-data-StormData.csv.bz2")
    bunzip2("repdata-data-StormData.csv.bz2", overwrite=T, remove=F)
}
```

### Load Variable
Load and investigate the data set.  Cache is turned on to improve performance.

```{r LoadData, cache=TRUE}
  if(!"StormData" %in% ls())
  {
    StormData<-read.csv("repdata-data-StormDatarepdata-data-StormData.csv", stringsAsFactors = FALSE)
  }
  dim(StormData)
  head(StormData, n=1)
  str(StormData)
```

## Investigate Data set
Histogram of storm events

```{r histStormEvents, cache=TRUE}
  StormData$Year<- as.numeric(year(mdy_hms(StormData$BGN_DATE)))
  hist(StormData$Year, breaks=61)

```
## Chosing the best subset of the data
It is unlikely that there were significantly more storms starting around 1990.  More likely record keeping improved; so the analysis will focus only on storms after 1990
```{r reshapeData}
  StormData<-StormData[StormData$Year >= 1990, ]
  dim(StormData)
```

## Build Summary Statistics
### Fatalities
```{r Fatalities}
  Fatalities<-aggregate(FATALITIES~EVTYPE, data=StormData, sum)
  Fatalities<-head(arrange(Fatalities, Fatalities$FATALITIES, decreasing=TRUE), n=10)
  Fatalities
  
```

### Injuries
```{r Injuries}
  Injuries<-aggregate(INJURIES~EVTYPE, data=StormData, sum)
  Injuries<-head(arrange(Injuries, Injuries$INJURIES, decreasing=TRUE), n=10)
  Injuries
  
```
